Using best practices for forecasting biodiversity changes in breeding birds

### 2017-07-12 manuscript/draft.Rmd
re-order the results

Bird species richness was obtained from the Breeding Bird Survey (BBS)
[@pardieck2017] using the Data Retriever software [@morris2013] and
rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
all nocturnal, cepuscular, and aquatic species [since these species are not well
sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
hybrids. Due to limitations of the environmental data (below), we did not
include any data from outside the continental United States. Observed richness
from 1982 (the first complete year of NDVI data) to 2003 were used for training
the models, and data from 2004 to 2013 was used for testing their performance.
Only BBS routes that were sampled during 70% of the years in the training period
were used (i.e., routes with at least 16 annual observations). All data from
routes that did not meet BBS acceptable criteria were excluded, as were routes
that had not been visited during at least 70% of the years in our model-training
window. The resulting dataset included [[N]] species and [[N]] sites. Among
sites,  richness varied from [[N]] to [[N]] with an average richness of [[N]]
species. Within-site richness variation (among years) was small (standard
deviation of about 5 species, after controlling for observer effects).

@@

The site-observer model found that 70% of the variance in richness in the
training set could be explained by differences among sites, and 21% could be
explained by differences among observers. The remaining 9% represents residual
variation, where a given observer might report a different number of species in
different years. There was little temporal autocorrelation in these residuals
([[about 0.04, +/- something]]), suggesting that richness is close to stationary
on these time scales once observer differences are accounted for.

When comparing forecasts for richness across sites all methods performed well
(Figure 2; R^2 XX-XX%). However stacked and joint SDMs and the macroecological
model both failed to successfully forecast the highest richness sites resulting
in a notable clustering of predicted values near ~60 species and the poorest
performance in the cross-site comparison (R^2 XX-XX%).

For most combinations of metric and forecast time lag, the best-performing model
for forecasting within-site variability was either the "average" baseline model
or the ARIMA model [[Not sure what the bit about within-site variability means
here. Also, didn't Naive usually win when observer effects weren't corrected
for?]] (Figure 3). In general, each model's error increased more-or-less
linearly with time, although the slopes and initial error rates varied across
models (Figure 3). The site-level models (i.e. the baselines and the ARIMA
methods) tended to do very well (Figure 3) on short time horizons, especially in
terms of RMSE.

In general, the models produced confidence intervals that were too narrow,
indicating overconfident predictions. Of these, the random forest-based SDM
stack was the most overconfident, with observed richness values frequently
falling well outside their confidence limits (Figure 3C). As noted elsewhere
[@harris2015], this overconfidence is a product of the assumption in stacked
SDMs that errors in the species-level predictions are independent. The JSDM and
macroecological model has the best coverage, though each model still
underestimated the possible variation in richness. The "naive" model
was the only model whose confidence intervals were too wide (Figure 3), which
can be attributed to the rapid rate at which these intervals expand as the time
lag increases (Figure 1).

RMSE and deviance produced qualitatively similar results in most cases, with the
notable exception of the stacked SDMs (Figure 3A and 3B). Deviance, unlike RMSE,
depends on the full predictive distribution of a model, including the model's
estimated uncertainty. As a result, the stacked SDMs' deviance was notably
higher than the next-worst model even though its RMSE was not unusual. On the
other hand, the GBM-based "macroecological" model and especially the mistnet
JSDM had reasonably well-calibrated uncertainty estimates, even in the presence
of uncorrected observer biases (Figure 3C). This prevented their deviance from
rising as quickly as the site-level models as the observer population evolved
and resulted in better comparative performance for these models in terms of
deviance (Figure 3B).

Accounting for the uncertainty associated with differences among observers
resulted in improvements to all metrics of fit (Figure 4). The one exception was
RMSE of the "Naive" model, which remained unchanged. The improvements in point
estimates primarily resulted from large improvements in a small number of
forecasts, presumably ones where a large difference between observers resulted
in a large shift in predicted richness values that the models could not
otherwise accommodate. Most forecasts exhibited only small changes in RMSE.
However these changes were sufficient to change the rank ordering of model
performance. For example, the "naive" model generally had the lowest average
error when observer effects were not explicitly accounted for, switching to the
"average" and ARIMA model once observer differences were corrected. Improvements
in coverage were notable for all models except Naive and JSDM, meaning that
incorporating observer effects resulted in improved quantification of
uncertainty. The lack of change in the mistnet JSDM seems to indicate that its
latent variables were already accounting for observer-related uncertainty. These
differences meant that, in terms of deviance, neither the JSDM nor the
macroecological model model benefitted as much from the as the site-level models
did from the observer corrections. As a result, the average and ARIMA models
maintained a small average advantage over the JSDM in terms of deviance at
longer time scales, rather than falling behind after about five years.

>>>>

Bird species richness was obtained from the Breeding Bird Survey (BBS)
[@pardieck2017] using the Data Retriever software [@morris2013] and
rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
all nocturnal, cepuscular, and aquatic species [since these species are not well
sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
hybrids. Due to limitations of the environmental data (below), we did not
include any data from outside the continental United States. Observed richness
from 1982 (the first complete year of NDVI data) to 2003 were used for training
the models, and data from 2004 to 2013 was used for testing their performance.
Only BBS routes that were sampled during 70% of the years in the training period
were used (i.e., routes with at least 16 annual observations). All data from
routes that did not meet BBS acceptable criteria were excluded, as were routes
that had not been visited during at least 70% of the years in our model-training
window. The resulting dataset included [[N]] species and [[N]] sites. Among
sites,  richness varied from [[N]] to [[N]] with an average richness of [[N]]
species.

@@

The site-observer model found that 70% of the variance in richness in the
training set could be explained by differences among sites, and 21% could be
explained by differences among observers. The remaining 9% represents residual
variation, where a given observer might report a different number of species in
different years. In the training set, the residuals had a standard deviation of
about [[+/- 5]] species. After correcting for observer differences, there was
little temporal autocorrelation in these residuals ([[about 0.04, +/-
something]]), suggesting that richness is approximately stationary on these time
scales.

When comparing forecasts for richness across sites all methods performed well
(Figure 2; R^2 XX-XX%). However SDMs (both stacked and joint) and the
macroecological model all failed to successfully forecast the highest-richness
sites resulting in a notable clustering of predicted values near ~60 species and
the poorest performance in the cross-site comparison (R^2 XX-XX%).

In general, each model's error increased more-or-less linearly with time,
although the slopes and initial error rates varied across models (Figure 3). The
site-level models (i.e. the baselines and the ARIMA methods) tended to do very
well (Figure 3), especially on short time horizons, and especially in terms of
RMSE. Of these, the "Naive" model performed especially well at sites with
uncorrected observer turnover (since its predictions were best-calibrated to
observers that were active in later years).

Most models produced confidence intervals that were too narrow, indicating
overconfident predictions. The random forest-based SDM stack was the most
overconfident model, with observed richness values frequently falling well
outside its confidence limits (Figure 3C). This stacked SDM's narrow
predictive distribution also caused it to have notably higher deviance (Figure
3B) than the next-worst model, even though its point estimates were not
unusually bad in terms of RMSE (3A), and the other models' relative performance
was not nearly as sensitive to the choice of error metric. As discussed
elsewhere [@harris2015], this overconfidence is a product of the assumption in
stacked SDMs that errors in the species-level predictions are independent.
Because they avoided this assumption, the GBM-based "macroecological" model and
especially the mistnet JSDM had reasonably well-calibrated uncertainty
estimates, even in the presence of uncorrected observer biases (Figure 3C). This
calibration improved both models' deviance relative to the baseline models,
especially as the observer population evolved (Figure 3B). The "naive" model was
the only model whose confidence intervals were too wide (Figure 3), which can be
attributed to the rapid rate at which these intervals expand as the time lag
increases (Figure 1).

Accounting for the uncertainty associated with differences among observers
resulted in improvements to nearly all metrics of fit (Figure 4). The
improvements in point estimates (as measured by RMSE) primarily resulted from
large improvements in a small number of forecasts where a large difference
between observers resulted in a large shift in predicted richness values that
the models could not otherwise accommodate. The "Average" and ARIMA models
generally showed the largest improvement here, and the Naive model showed the
smallest (since it was already very responsive to observer turnover). As a
result, the "Average" and "ARIMA" models tended to perform the best, once
observer differences were accounted for. Improvements in coverage were notable
for all models except Naive and JSDM, meaning that unmodeled observer effects
had been partly responsible for the poor quantification of uncertainty. As a
result, the average and ARIMA models maintained a small average advantage over
the JSDM in terms of deviance at longer time scales, rather than falling behind
after about five years as in Figure 3B.
